---
title: "**Introduction to scatterplots with ggplot2**"
author: "Richard Layton"
date: "2017-09-04"
output: beamer_presentation
---

```{r setup, include=FALSE}
library(knitr)
opts_knit$set(root.dir = "../")
opts_chunk$set(echo=TRUE, message=FALSE, collapse=FALSE)
opts_chunk$set(fig.height = 5.5)
options(width = 52)
library(tidyverse)
# set pander table-layout options
library(pander)
panderOptions('table.alignment.default', function(df)
    ifelse(sapply(df, is.numeric), 'right', 'left'))
panderOptions('table.split.table', Inf)
panderOptions('big.mark', ",")
panderOptions('keep.trailing.zeros', TRUE)
```

## R package vocabulary

```{r echo=FALSE}
df <- tribble(
	~Term, ~Definition,
	"package", "An R 'app', the basic unit of shareable code, a collection of R functions, data, and code",
	"library", "The directory on your computer where packages are stored, e.g., C:/R/library",
	"CRAN", "Online home of 11,340 R packages (2017-09-02), the Comprehensive R Archive Network",
	"install", "Download a package, save it in your library",
	"update", "Bring the package in your library up to date",
	"load", "To use a package in an R session, you load the package using the library() command."
)
```

```{r echo=FALSE}
pander(df, split.cells = c(30, 60))
```

## Managing packages in RStudio

\includegraphics{../images/panes-5-packages.png}

## The graphics package we’ll use most often is ggplot2 

\begin{columns}[T]

\begin{column}[T]{0.40\textwidth}

The Tidyverse: R packages for data science

\includegraphics{../images/tidyverse.png}

\end{column}

\begin{column}[T]{0.60\textwidth}

When you get a chance, install the {\tt tidyverse} package

\vspace{0.5\baselineskip}
{\tt tidyverse} includes {\tt ggplot2} and many other packages we will find useful 

\vspace{0.5\baselineskip}
\begin{itemize}
\item {\tt ggplot2} is the package
\item {\tt ggplot()} is the function
\end{itemize}

\end{column}

\end{columns}

## The arguments of `ggplot()`

```{r echo=FALSE}
df <- tribble(
	~Arguments, ~Definition,
	"data", "a data frame",
	"aesthetic", "visual properties: color, shape, etc.",
	"geom", "geometric shape: points, lines, rectangles, etc.",
	"scale", "map data units to computer units",
	"stat", "summarize or transform data, e.g., a regression",
	"facet", "create small multiples"
)
pander(df, split.cells = c(30, 60))
```

```{r echo=FALSE}
rm(df)
```

## Example

Getting the example started 

- I installed the `VGAMdata` package from CRAN

- Open a new R script

- Load the package into my workspace using `library()`

```{r}
library(VGAMdata)
```

- The 2012 summer Olympic data about individual athletes is accessed using the `data()` function

```{r}
data("oly12")
```


## What objects?

What objects are in my workspace?

```{r}
ls()
```

What class of object is it?

```{r}
class(oly12)
```

Good. It's a data frame. 

## Examine the data frame

What variables (columns) do I have?

```{r}
names(oly12)
```

To see the structure of the data frame in more detail, use 

```{r results='hide'}
glimpse(oly12)
```


## 

```{r echo=FALSE}
glimpse(oly12)
```


## Data and aesthetics

```{r}
ggplot(data = oly12, aes(x = Height, y = Weight))
```


## Add points

```{r}
ggplot(data = oly12, aes(x = Height, y = Weight)) +
	geom_point()
```

## Select 4 columns and filter for complete cases only  

```{r}
oly12_sub <- oly12 %>%
	select(Height, Weight, Sex, Sport) %>% 
	filter(complete.cases(.))
glimpse(oly12_sub)
```

## Graph it again (no waning message)

```{r}
ggplot(data = oly12_sub, aes(x = Height, y = Weight)) +
	geom_point()
```

## Jitter the data markers to reduce overprinting

```{r}
ggplot(data = oly12_sub, aes(x = Height, y = Weight)) +
	geom_jitter()
```

## Edit the transparency of the data markers 

```{r}
ggplot(data = oly12_sub, aes(x = Height, y = Weight)) +
	geom_jitter(alpha = 0.2)
```

## Edit the axis labels

```{r}
ggplot(data = oly12_sub, aes(x = Height, y = Weight)) +
	geom_jitter(alpha = 0.2) +
	labs(x = "Height (m)", y = "Weight (kg)")
```


## Bivariate? Trivariate?    


- A scatterplot displays two continuous variables
- How can a scatterplot be tri-variate? 

```{r echo=FALSE}
glimpse(oly12_sub)
```


## Add a facet by Sex

```{r}
ggplot(data = oly12_sub, aes(x = Height, y = Weight)) +
	geom_jitter(alpha = 0.2) +
	labs(x = "Height (m)", y = "Weight (kg)") +
	facet_wrap(~Sex)
```

## Or by Sport, creating a small-multiples graph

```{r eval=FALSE}
ggplot(data = oly12_sub, aes(x = Height, y = Weight)) +
	geom_jitter(alpha = 0.2) +
	labs(x = "Height (m)", y = "Weight (kg)") +
	facet_wrap(~Sport)
```

##

```{r echo = FALSE, fig.height = 7}
ggplot(data = oly12_sub, aes(x = Height, y = Weight)) +
	geom_jitter(alpha = 0.2) +
	labs(x = "Height (m)", y = "Weight (kg)") +
	facet_wrap(~Sport)
```



## We can include a fourth variable using a color aesthetic

```{r eval=FALSE}
ggplot(data = oly12_sub, 
	  aes(x = Height, y = Weight, col = Sex)) +
	geom_jitter(alpha = 0.2) +
	labs(x = "Height (m)", y = "Weight (kg)") +
	facet_wrap(~Sport)
```

##

```{r echo = FALSE, fig.height = 7}
ggplot(data = oly12_sub, 
		aes(x = Height, y = Weight, col = Sex)) +
	geom_jitter(alpha = 0.2) +
	labs(x = "Height (m)", y = "Weight (kg)") +
	facet_wrap(~Sport)
```



## Review

\begin{columns}[T]

\begin{column}[T]{0.33\textwidth}

Packages

\begin{itemize}
\item tidyverse
\item VGAMdata
\end{itemize}

\vspace{0.5\baselineskip}
Tidyverse functions

\begin{itemize}
\item ggplot()
\item glimpse()
\item select()
\item filter()
\end{itemize}

\end{column}

\begin{column}[T]{0.33\textwidth}

R functions

\begin{itemize}
\item library()
\item data()
\item ls()
\item class()
\item names()
\item print()
\item complete.cases()
\end{itemize}

\end{column}

\begin{column}[T]{0.33\textwidth}

ggplot arguments

\begin{itemize}
\item aes()
\item geom\_point()
\item geom\_jitter()
\item labs()
\item facet\_wrap()
\end{itemize}

\end{column}

\end{columns}














