---
title: "**Data basics**"
author: "Richard Layton"
date: "2017-09-05"
output: beamer_presentation
---

```{r setup, include=FALSE}
library(knitr)
opts_knit$set(root.dir = "../")
opts_chunk$set(echo=TRUE, message=FALSE, collapse=FALSE)
opts_chunk$set(fig.height = 5.5)
options(width = 52)
```


```{r include=FALSE}
library(tidyverse)
library(readxl)
library(pander)
panderOptions('table.alignment.default', function(df)
    ifelse(sapply(df, is.numeric), 'right', 'left'))
panderOptions('table.split.table', Inf)
panderOptions('big.mark', ",")
panderOptions('keep.trailing.zeros', TRUE)
```

## Today's R packages 

```{r echo=FALSE}
df <- tribble(
	~Package, ~For,
	"tidyverse", "ggplot2, dplyr, readr, tidyr", 
	"ggplot2",   "Graphing data", 
	"dplyr",     "Data manipulation", 
	"readr",     "Read rectangular data files (like csv or tsv)", 
	"tidyr",     "set of functions that help you get to tidy data", 
	"readxl",    "Read .xls and .xslx files",
	"VIM",       "Examine missing values in a data frame"
)
```

```{r echo=FALSE}
pander(df, split.cells = c(30, 60))
```

## Learning objectives 

After working through the data basics tutorial, you should be able to

- Read data from an Excel file 
- Read and write CSV data files 
- Obtain and manipulate data sets included with R and R packages 

\begin{center}
\includegraphics{../images/data-basics-excel-to-csv.png}
\end{center}





## The readxl package

\vspace{0.5\baselineskip}
```{r eval=FALSE}
library(readxl)
my_data <- read_excel(
       path = "data/my_file_name.xlsx"
    , sheet = "sheet_tab_name"
    )
```

- To read .xls or .xlsx files
- Assumes data is tidy or nearly tidy

\vspace{0.5\baselineskip}
\begin{center}
\includegraphics[scale=0.6]{../images/data-basics-excel-long.png}
\end{center}

## What is tidy data?

\vspace{0.5\baselineskip}
\begin{center}
\includegraphics[scale=0.6]{../images/data-basics-tidy-schematic.png}
\end{center}

\vspace{5\baselineskip}
{\footnotesize
Source: data-wrangling-cheatsheet,  https://www.rstudio.com/wp-content/uploads/2015/02/data-wrangling-cheatsheet.pdf
}

## When your data is not tidy

\vspace{0.5\baselineskip}
\begin{center}
\includegraphics[scale=0.5]{../images/data-basics-excel-not-tidy.png}
\end{center}




## When your data is not tidy

```{r}
wide_data <- read_excel("data/VADeaths.xlsx")
```

```{r echo=FALSE}
kable(wide_data)
```



## We can skip lines, but might lose information 

\vspace{0.5\baselineskip}
```{r}
wide_data <- read_excel("data/VADeaths.xlsx", skip = 1)
```

```{r echo=FALSE}
kable(wide_data)
```

Reading and reshaping data like this is a topic for another day 
 











<!-- ## Managing packages in RStudio -->

<!-- \includegraphics{../images/panes-5-packages.png} -->

<!-- ## The graphics package we’ll use most often is ggplot2  -->

<!-- \begin{columns}[T] -->

<!-- \begin{column}[T]{0.40\textwidth} -->

<!-- The Tidyverse: R packages for data science -->

<!-- \includegraphics{../images/tidyverse.png} -->

<!-- \end{column} -->

<!-- \begin{column}[T]{0.60\textwidth} -->

<!-- When you get a chance, install the {\tt tidyverse} package -->

<!-- \vspace{0.5\baselineskip} -->
<!-- {\tt tidyverse} includes {\tt ggplot2} and many other packages we will find useful  -->

<!-- \vspace{0.5\baselineskip} -->
<!-- \begin{itemize} -->
<!-- \item {\tt ggplot2} is the package -->
<!-- \item {\tt ggplot()} is the function -->
<!-- \end{itemize} -->

<!-- \end{column} -->

<!-- \end{columns} -->

<!-- ## The arguments of `ggplot()` -->

<!-- ```{r echo=FALSE} -->
<!-- df <- tribble( -->
<!-- 	~Arguments, ~Definition, -->
<!-- 	"data", "a data frame", -->
<!-- 	"aesthetic", "visual properties: color, shape, etc.", -->
<!-- 	"geom", "geometric shape: points, lines, rectangles, etc.", -->
<!-- 	"scale", "map data units to computer units", -->
<!-- 	"stat", "summarize or transform data, e.g., a regression", -->
<!-- 	"facet", "create small multiples" -->
<!-- ) -->
<!-- pander(df, split.cells = c(30, 60)) -->
<!-- ``` -->

<!-- ```{r echo=FALSE} -->
<!-- rm(df) -->
<!-- ``` -->

<!-- ## Example -->

<!-- Getting the example started  -->

<!-- - I installed the `VGAMdata` package from CRAN -->

<!-- - Open a new R script -->

<!-- - Load the package into my workspace using `library()` -->

<!-- ```{r} -->
<!-- library(VGAMdata) -->
<!-- ``` -->

<!-- - The 2012 summer Olympic data about individual athletes is accessed using the `data()` function -->

<!-- ```{r} -->
<!-- data("oly12") -->
<!-- ``` -->


<!-- ## What objects? -->

<!-- What objects are in my workspace? -->

<!-- ```{r} -->
<!-- ls() -->
<!-- ``` -->

<!-- What class of object is it? -->

<!-- ```{r} -->
<!-- class(oly12) -->
<!-- ``` -->

<!-- Good. It's a data frame.  -->

<!-- ## Examine the data frame -->

<!-- What variables (columns) do I have? -->

<!-- ```{r} -->
<!-- names(oly12) -->
<!-- ``` -->

<!-- To see the structure of the data frame in more detail, use  -->

<!-- ```{r results='hide'} -->
<!-- glimpse(oly12) -->
<!-- ``` -->


<!-- ##  -->

<!-- ```{r echo=FALSE} -->
<!-- glimpse(oly12) -->
<!-- ``` -->


<!-- ## Data and aesthetics -->

<!-- ```{r} -->
<!-- ggplot(data = oly12, aes(x = Height, y = Weight)) -->
<!-- ``` -->


<!-- ## Add points -->

<!-- ```{r} -->
<!-- ggplot(data = oly12, aes(x = Height, y = Weight)) + -->
<!-- 	geom_point() -->
<!-- ``` -->

<!-- ## Select 4 columns and filter for complete cases only   -->

<!-- ```{r} -->
<!-- oly12_sub <- oly12 %>% -->
<!-- 	select(Height, Weight, Sex, Sport) %>%  -->
<!-- 	filter(complete.cases(.)) -->
<!-- glimpse(oly12_sub) -->
<!-- ``` -->

<!-- ## Graph it again (no warning message) -->

<!-- ```{r} -->
<!-- ggplot(data = oly12_sub, aes(x = Height, y = Weight)) + -->
<!-- 	geom_point() -->
<!-- ``` -->

<!-- ## Jitter the data markers to reduce overprinting -->

<!-- ```{r} -->
<!-- ggplot(data = oly12_sub, aes(x = Height, y = Weight)) + -->
<!-- 	geom_jitter() -->
<!-- ``` -->

<!-- ## Edit the transparency of the data markers  -->

<!-- ```{r} -->
<!-- ggplot(data = oly12_sub, aes(x = Height, y = Weight)) + -->
<!-- 	geom_jitter(alpha = 0.2) -->
<!-- ``` -->

<!-- ## Edit the axis labels -->

<!-- ```{r} -->
<!-- ggplot(data = oly12_sub, aes(x = Height, y = Weight)) + -->
<!-- 	geom_jitter(alpha = 0.2) + -->
<!-- 	labs(x = "Height (m)", y = "Weight (kg)") -->
<!-- ``` -->


<!-- ## Bivariate? Trivariate?     -->


<!-- - A scatterplot displays two continuous variables -->
<!-- - How can a scatterplot be tri-variate?  -->

<!-- ```{r echo=FALSE} -->
<!-- glimpse(oly12_sub) -->
<!-- ``` -->


<!-- ## Add a facet by Sex -->

<!-- ```{r} -->
<!-- ggplot(data = oly12_sub, aes(x = Height, y = Weight)) + -->
<!-- 	geom_jitter(alpha = 0.2) + -->
<!-- 	labs(x = "Height (m)", y = "Weight (kg)") + -->
<!-- 	facet_wrap(~Sex) -->
<!-- ``` -->

<!-- ## Or by Sport, creating a small-multiples graph -->

<!-- ```{r eval=FALSE} -->
<!-- ggplot(data = oly12_sub, aes(x = Height, y = Weight)) + -->
<!-- 	geom_jitter(alpha = 0.2) + -->
<!-- 	labs(x = "Height (m)", y = "Weight (kg)") + -->
<!-- 	facet_wrap(~Sport) -->
<!-- ``` -->

<!-- ## -->

<!-- ```{r echo = FALSE, fig.height = 7} -->
<!-- ggplot(data = oly12_sub, aes(x = Height, y = Weight)) + -->
<!-- 	geom_jitter(alpha = 0.2) + -->
<!-- 	labs(x = "Height (m)", y = "Weight (kg)") + -->
<!-- 	facet_wrap(~Sport) -->
<!-- ``` -->



<!-- ## We can include a fourth variable using a color aesthetic -->

<!-- ```{r eval=FALSE} -->
<!-- ggplot(data = oly12_sub,  -->
<!-- 	  aes(x = Height, y = Weight, col = Sex)) + -->
<!-- 	geom_jitter(alpha = 0.2) + -->
<!-- 	labs(x = "Height (m)", y = "Weight (kg)") + -->
<!-- 	facet_wrap(~Sport) -->
<!-- ``` -->

<!-- ## -->

<!-- ```{r echo = FALSE, fig.height = 7} -->
<!-- ggplot(data = oly12_sub,  -->
<!-- 		aes(x = Height, y = Weight, col = Sex)) + -->
<!-- 	geom_jitter(alpha = 0.2) + -->
<!-- 	labs(x = "Height (m)", y = "Weight (kg)") + -->
<!-- 	facet_wrap(~Sport) -->
<!-- ``` -->



<!-- ## Review -->

<!-- \begin{columns}[T] -->

<!-- \begin{column}[T]{0.33\textwidth} -->

<!-- Packages -->

<!-- \begin{itemize} -->
<!-- \item tidyverse -->
<!-- \item VGAMdata -->
<!-- \end{itemize} -->

<!-- \vspace{0.5\baselineskip} -->
<!-- Tidyverse functions -->

<!-- \begin{itemize} -->
<!-- \item ggplot() -->
<!-- \item glimpse() -->
<!-- \item select() -->
<!-- \item filter() -->
<!-- \end{itemize} -->

<!-- \end{column} -->

<!-- \begin{column}[T]{0.33\textwidth} -->

<!-- R functions -->

<!-- \begin{itemize} -->
<!-- \item library() -->
<!-- \item data() -->
<!-- \item ls() -->
<!-- \item class() -->
<!-- \item names() -->
<!-- \item print() -->
<!-- \item complete.cases() -->
<!-- \end{itemize} -->

<!-- \end{column} -->

<!-- \begin{column}[T]{0.33\textwidth} -->

<!-- ggplot arguments -->

<!-- \begin{itemize} -->
<!-- \item aes() -->
<!-- \item geom\_point() -->
<!-- \item geom\_jitter() -->
<!-- \item labs() -->
<!-- \item facet\_wrap() -->
<!-- \end{itemize} -->

<!-- \end{column} -->

<!-- \end{columns} -->














