---
title: "What I hope you take away"
author: "Richard Layton"
date: "2017-11-10"
output: beamer_presentation
---

```{r setup, include=FALSE}
library(knitr)
opts_knit$set(root.dir = "../")
opts_chunk$set(echo = TRUE, message = FALSE, collapse = FALSE)
opts_chunk$set(fig.height = 5.5)
options(width = 52)
```

```{r include=FALSE}
library(tibble)
library(readr)
library(readxl)
library(pander)
panderOptions('table.alignment.default', function(df)
    ifelse(sapply(df, is.numeric), 'right', 'left'))
panderOptions('table.split.table', Inf)
panderOptions('big.mark', ",")
panderOptions('keep.trailing.zeros', TRUE)
```

## The big ideas

To manage your project, manage your files

Understand your data  

Explore stories with visualizations    

Be guided by the rhetorical situation 

Design the final document   


## To manage your project, manage your files

- Establish an organization scheme and stick to it

\smallskip

- Keep scripts short (< 100 lines) and do one thing 

\smallskip

- `data`  directory for data files and R scripts that tidy them 

\smallskip

- `design` directory for scripts to turn data into visuals

\smallskip

- `visuals` directory to receive the graphs 

\smallskip

- `portfolio` R Markdown script to assemble the document  


 





## Understand your data 

- Acquire interesting data [(recommended links)](https://github.com/DSR-RHIT/me447-visualizing-data/blob/master/cm/tut-1202_data-links.md)

\smallskip

- Avoid surface learning: seek to understand your data structure  

\smallskip

- Develop the necessary skills and make the data tidy 

\smallskip

- Seek additional data as warranted by the rhetorical situation 






   

 

## Explore stories with visualizations 

- Use the data structure to inform design sketches (no code) 

\smallskip

- Learn from prior art,  use small data sets you understand

\smallskip 

- Learn from prior art, understand one line of code at a time   

\smallskip

- One tidy data set, 15 visualizations 

\smallskip

- Let the rhetorical situation add to your exploration 



 



## Be guided by the rhetorical situation 


- Compared to what? Larger context? Who cares? 


\smallskip 

- Understand your audience 

\smallskip

- Judiciously apply relevant design principles

\smallskip

- Tell the important story for that audience 

\smallskip

- Expect the data graphic to raise new question. Seek answers. 

\smallskip

- Repeat




## Design your final document


**Design** means making deliberate choices: 

\smallskip

- Every pixel in every display

\smallskip

- Every word, sentence, and paragraph

\smallskip

- Every element of document design: typefaces, spacing, etc. 



