---
output: github_document
bibliography: "../resources/bib/refs.bib" 
csl: "../resources/styles/journal-of-peace-research.csl"
link-citations: yes
---

# portfolio data

```{r echo=FALSE}
library(knitr)
opts_knit$set(root.dir = "../")
opts_chunk$set(echo = FALSE, message = FALSE, collapse = TRUE, warning = FALSE)
```

Generally, different displays in the portfolio have different structures with different numbers of quantitative and categorical variables. The basic data structure for all graphs in the portfolio are summarized in the table. 


```{r}
library(tidyverse)
df <- tribble(
	~display, ~type, ~quantitative, ~categorical, 
	"[D1 Distributions](cm210_d1-distributions.md)", 
	    "box plot",      "one", "one, min. 3 levels", 
	"", "histogram",     "one", "one, min. 3 levels",
	"", "density plot",  "one", "one, min. 3 levels",
	
	"D2 Multiway", 
	"multiway", "one", "two, min. 3 levels each", 
	
	"D3 Correlations", 
	"scatterplot", "two", "one, min. 3 levels", 
	
	"D4 Injuries or fatalities", 
	      "dot plot",   "one", "optional", 
  "",   "line graph", "two", "optional", 
	
	"D5 Fix a lie", 
	      "dot plot",    "one", "optional", 
	"",   "line graph",  "two", "optional", 
	"",   "scatterplot", "two", "optional", 
	
	"D6 Hypervariate", 
	    "conditioning plot",   "three or four", "", 
	"", "scatterplot matrix",  "four or more", "", 
	"", "parallel coordinate", "four or more", "optional", 
	
	"D7 Counts", 
		  "q-q",            "one", "two distributions", 
	"", "cycle plot",     "one", "two in time series, e.g., month & year", 
	"", "Sankey diagram", "one", "N nodes dividing the flow",
	"", "mosaic",         "one", "three or more, min. 2 levels each", 
	"", "diverging stacked bar", "one", "survey questions, min. 5" , 
	"D7 Spatial data", "quantitative map", "one", "", 
	"",        "qualitative map", "", "one, N levels"
	)
kable(df)
```




<!-- ## distributions {#distributions} -->

<!-- [portfolio data] -->



## notes

1. *Quantitative variables are typically continuous*. If you encounter a quantitative variable (other than time) that is discrete rather than continuous, please consult with me on its suitability.

1. *Is a categorical variable ordered?* The order in which the levels of a category are presented depends on whether or not the category is ordered or not.

1. *Some categorical variables are potentially optional*. A category marked   "optional" indicates that the graph may or may not have a categorical variable whose levels are used as a conditioning variable.

1. *D2 Multiway data*. The distinguishing characteristic of multiway data is that there is a quantitative value associated with *every combination* of levels of the two categorical variables. Also, draft both multiway duals and select  the one that best conveys the stories in the data.

1. *D4 Injuries or fatalities displayed ethically*. The data must convey information about human injury or fatality (similar to the data in *Cruel Pies*). Consult with me on your data selection---the graph type depends on the data. The two graph types shown are the most commonly encountered.

1. *D5 Fix a lie*. Consult with me on your data selection---the graph type depends on the data. The two graph types shown are the most commonly encountered.

1. *D7* is a graph type you learn on your own. The two general categories are *Counts* or the frequency with which something occurs and *Spatial data* that are displayed on maps.




