---
output: github_document
bibliography: "../resources/bib/refs.bib" 
csl: "../resources/styles/journal-of-peace-research.csl"
link-citations: yes
---

# portfolio display requirements

```{r echo=FALSE}
library(knitr)
opts_knit$set(root.dir = "../")
opts_chunk$set(echo = FALSE, message = FALSE, collapse = TRUE, warning = FALSE)
```

The table summarizes the basic data structure for all portfolio graphs.    
Specific data requirements are provided below.     

<!-- Quantitative variables are typically continuous. If you encounter a quantitative variable (other than time) that is discrete rather than continuous, please consult with me on its suitability for your graph.   -->

```{r}
library(tidyverse)
df <- tribble(
	~display, ~type, ~quantitative, ~categorical, 
	"[D1 distributions]", "box plot, histogram, or density plot", "one", "one",
	"[D2 multiway]", "multiway", "one", "two", 
	"[D3 correlations]", "scatterplot", "two", "one", 
	"[D4 injuries or fatalities]", "dot plot", "one", "optional", 
  "", "line graph or scatterplot", "two", "optional",
	"[D5 redesign a graphical lie]", "dot plot", "one", "optional", 
	"", "line graph or scatterplot", "two", "optional",
	"[D6 multivariate]", "conditioning plot", "three or four", "", 
	"", "scatterplot matrix or parallel coordinate", "four or more", "", 
	"[D7 self-taught]", "q-q", "one", "one", 
	"", "cycle plot", "one", "two", 
	"", "Sankey diagram", "one", "two or more",
	"", "mosaic", "one", "three or more", 
	"", "diverging stacked bar", "one", "one" , 
	"", "quantitative map", "one", "", 
	"", "qualitative map", "", "one"
	)
kable(df)
```



## D1 distributions 

Univariate data are measurements of a single quantitative variable. The data can be separated into groups by a categorical variable, creating one distribution for each level of the category for comparison. 

We will use three prospective graph designs for comparing distributions. Include only one of these graph types in your portfolio. Select the design that best conveys the stories in your data. 

- histograms 
- density plots 
- box plots 

Data requirements

- Minimum 200 observations of one quantitative variable  
- One or two categorical variables  
- No time-series 

Categorical variables 

- If one category is used, it must have at least three levels for a minimum of three comparisons.  
- If two categories are used, they must have at least two levels each for a minmum of four comparisons. 

[Portfolio display requirements]



## D2 multiway 

The distinguishing characteristic of multiway data is that there is a quantitative value associated with every combination of levels of two categorical variables. 

Graph both the original multiway and its dual (swapping rows and panels). Include only one of these graphs in your portfolio. Select the design that best conveys the stories in your data. 

Data requirements

- One quantitative variable with a minimum of 24 observations
- Two categorical variables with minimum numbers of levels as follows
  - 3 levels in category 1 and at least 8 levels in category 2 
  - 4 levels in category 1 and at least 6 levels in category 2   
  - 5 levels in category 1 and at least 5 levels in category 2 

[Portfolio display requirements]



## D3 correlations



[Portfolio display requirements]



## D4 injuries or fatalities

The data must convey information about human injury or fatality (similar to the data in *Cruel Pies*). Consult with me on your data selection---the graph type depends on the data. The two graph types shown are the most commonly encountered.

[Portfolio display requirements]



## D5 redesign a graphical lie

Consult with me on your data selection---the graph type depends on the data. The two graph types shown are the most commonly encountered.

[Portfolio display requirements]



## D6 multivariate

We use the term *multivariate* for data comprising 3 or more quantitative variables per observation.  

We use three graph designs for exploring multivariate data.

- conditioning plot 
- scatterplot matrix 
- parallel coordinate plot

Include only one of these graph types in your portfolio. Select the design that best conveys the stories in your data. 

Data requirements

- Minimum 20 observations  
- Three or more quantitative variables  
- No time-series data  
- No categorical variables  

Use a conditioning plot (coplot) for three or four variables. 

For four or more variables, use a scatterplot matrix or a parallel coordinate plot. 

[Portfolio display requirements]



## D7 self-taught

This is a graph type you learn to construct on your own. Select one only from the list and include it in the portfolio. 

Counts 

- [q-q plot]  
- [cycle plot]  
- [Sankey diagram]  
- [mosaic]  
- [diverging stacked bar]  

Spatial data 

- [quantitative map]  
- [qualitative map]  




## q-q plot

The quantile-quantile (q-q) plot is a graphical technique for determining if two groups of data come from populations with a common distribution. Data requirements: 

- Two samples of one quantitative variable, min. 200 observations total. 
- One categorical variable that defines the two groups (two levels only) 
- The sample sizes of the two groups do not need to be equal 
  
For additional technical details, see @nist-q-q. 




## cycle plot 



For additional technical details, see 



## Sankey diagram	




For additional technical details, see 


## mosaic	




## diverging stacked bar 

Diverging stacked bar charts are the primary graphical display technique for reporting survey data. Respondents are typically asked to choose one response from a discrete number of options such as Strongly Disagree, Disagree, No Opinion, Agree, Strongly Agree. 



For additional technical details, see @HeibergerRobbins2014. 





## quantitative map	


For additional technical details, see 




## qualitative map 

For additional technical details, see 






[Portfolio display requirements]



## references

<div id="refs"></div>

---  
[Calendar](../README.md#calendar)    
[Index](../README.md#index) 

