---
output: github_document
---

```{r setup, echo = FALSE}
library(knitr)
opts_knit$set(root.dir = "../")
opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, collapse = FALSE)
library(knitr)
library(readxl)
```

## portfolio requirements

An exceptional portfolio showcases your ability to communicate truthful and compelling stories using quantitative data and effective visual rhetoric. My goal is to help you craft a document that you would be proud to include in your employment portfolio. 

All coursework is directed towards completing a portfolio that demonstrates the degree to which you can: 

- Design effective and truthful data displays. 
- Credibly explain your design rationale. 
- Produce publication-quality visuals. 
- Credibly critique a data display. 

A portfolio checklist is included in your course handbook (or you can view it [here](../cm/folio-02_portfolio-checklist.pdf)). Use it to check off requirements as you complete them.

## the minimum 8 displays 

D1 through D8 are the minimum set of required displays. Lectures and tutorials are devoted to developing your ability to create each type.  

```{r echo = FALSE, results = 'hide'}
df <- read_excel("manage/04-data-portfolio-requirements.xlsx")
```

```{r echo = FALSE}
knitr::kable(df)
```

Notes on D7 and D8

1. You are to create your own designs and/or learn new display types on your own.  
1. You may create two different "learn on your own"" displays, two different original designs, or one of each. 
1. The text illustrates many display types we do not cover in class, for example, strip plots, month plots (cycle plots), co-plots, mosaics, and scatterplot matrices. Ideas are abundant online.

A display can meet more than one requirement. For example, creating a small multiples graph to satisfy the D6 requirement might also satisfy the requirements to have a graph in your technical discipline and be a redesign as well. 

Students often exceed the minimum number of displays to meet all the requirements, to explore data they find interesting, or to showcase something they've learned on their own.  




## written critiques

A written critique is required for every display. 

Your critique explains your design rationale. Use evidence and cite relevant principles of rhetoric, human perception, and design.

- The readings provide credible support to your critique arguments. 
- You present some of your displays in class to improve the displays and develop ideas for the critiques. 
- Plan to pre-write (elements of your argument), draft (it won’t be pretty but get it down), revise (writing is rewriting), edit (to polish), and check (facts and sources must be correct). 



## technical discipline data

Four of your displays must use data relevant to problems in your technical area. You might consider:

- Redesigning a poorly designed graph from a journal or conference proceedings in your technical area.
- Using data from any research you have been involved with.
- Obtaining data from a Rose faculty member who needs help with data graphics. 
- If you are a member of a team in another class or project you may act as a graph consultant for that team, creating data displays that contribute to the teams' work as well as your portfolio.  


## redesigns

A minimum of 4  displays must be redesigns of someone else's published display. At least one each of the following is required.

1. Redesign a multivariate bar chart or pie chart (a good candidate for D.3). 
2. Redesign a marketing infographic ([infographic examples](https://www.customermagnetism.com/infographics/what-is-an-infographic/)).
3. Redesign a multivariate line graph (a good candidate for D.4). 
4. Redesign a graphical lie. 

You may redesign to a type you've learned, e.g., converting a pie chart into a dot plot, or to an original design (a good candidate for D.7).

In the portfolio, show the original display and your redesign and critique both.

## a compelling story

At least one display must tell a compelling story

To compel an audience to action, a display must tell an important story based on complete and credible data and some insights into the story's context. You will encounter several such displays during the term: examples to inspire your work. This isn't easy, so only one is required. (A reproduction, even if compelling, does not satisfy this requirement.)

## the 500-level course

Students enrolled in the 500-level course must also use data from their project or thesis research to develop additional displays:

1. Three different types of exploratory displays for the same data, with critiques 
2. A final display suitable for publication in their thesis or project report, with critique 

## reproductions are optional

A maximum of 4 displays can be reproductions: designs you have copied from someone else. (The "someone else" may not be current or past Rose-Hulman faculty, staff, or students.) Reproductions come in two flavors: *imitations* and *clones.*

*Imitations.* Copy another's R code to display your data. Your data should have the same structure as theirs but dissimilar categories and observations. Adapt their code to your data and revise the reproduction to correct any major deficiencies. 

*Clones.* Write original code to copy another’s display. Find a published display you like but for which no code is available. You create original code to exactly recreate the display. Revise the reproduction to correct any major deficiencies. 

In the portfolio, show the original display and your reproduction and critique both.

## document design specifications

The portfolio is a single electronic document with a cover page and a table of contents. We will use the R Markdown feature in RStudio to combine text and R code in a single set of files that create a formatted MSWord document. 

- Typefaces: 10-12 pt with serifs for the main text. Headings, subheadings, and captions are your choice. 
- Margins 1 inch min., line spacing 1.15-1.25. This is an e-document, so page dimensions are not specified. US customary $8.5\times 11$ is acceptable but not required. 
- Colors unambiguous for color-vision-deficient (CVD) readers. 
- White space and gridding used effectively. 
- Page layout designed for the reader. 

*Page design*. Naomi Robbins' book illustrates an approach to page layout designed for the reader --- graphs and their critiques are on facing pages. A reader can read the text and see the graph it discusses without flipping pages.

## citing sources

Effective critiques are well-reasoned and well-informed. In our round-table critiques, you share and refine your reasoning; in our readings and discussions, you find a basis for an informed opinion. The credibility of your critique is supported by the credibility and relevance of your citations.

Avoid the novice practice of sprinkling citations like confetti at New Year's to make an argument *appear* supported. Instead, construct your argument first---say what you are trying to say---based on the stuff you're learning. That stuff is based on credible sources. Cite those sources.

We will use the author-date citation style and automate the generation of formatted citations and references using R Markdown. More on that later in the term.

Your portfolio concludes with a single bibliography. 


## first drafts $\neq$ final drafts

Drafting, revising, editing, and checking apply to the data displays as well as the critiques. 

As your knowledge and ability develop over the term, you may find that displays you created in the first few weeks no longer meet your developing standards of excellence. You may come across more interesting data or more compelling stories and decide to replace an earlier display with a new one.  

That's all to be expected with our goal of crafting a showcase document. 



---

[main page](../README.md)<br> 
[topics page]../(README-by-topic.md)
