---
output: github_document
bibliography: "../manage/refs.bib" 
csl: "../manage/csl/journal-of-peace-research.csl"
---

```{r echo=FALSE}
library(knitr)
opts_knit$set(root.dir = "../")
opts_chunk$set(echo=TRUE, message=FALSE, collapse=TRUE, warning=FALSE)
```

```{r echo=FALSE}
rm(list = (ls()))
library(tidyverse)
library(readxl)
```

## add a regression to a scatterplot

In the previous tutorial, we saved the pressure calibration data to a CSV file. 

```{r}
# read a CSV file
press_calibr <- read_csv("data/press-calibr-long.csv")
```

Create a scatterplot with the input pressure in psi on the x-axis and the senosr reading in psi on the y-axis. 

```{r}
# plot the calibration data in a scatterplot
f1 <- ggplot(data = press_calibr,
		aes(x = inputPSI, y = readingPSI)) +
	geom_point()
print(f1)
```

WE add a regression line (least-squares) by adding the *geom_smooth(method = lm)* function.  

```{r}
# add a linear regression
f2 <- f1 + geom_smooth(method = lm)
print(f2)
```

The gray band is the confidence interval. We can moit it using the *se = FALSE* argument. 

```{r}
# omit the confidence interval
f3 <- f1 +
	geom_smooth(method = lm, se = FALSE)
print(f3)
```

Next tutorial: [accessing data in R](tut-0407_access-data-in-R.md)

---

[main page](../README.md)<br> 
[topics page](../README-by-topic.md)


