---
output: github_document
bibliography: "../manage/refs.bib" 
csl: "../manage/csl/journal-of-peace-research.csl"
---

```{r echo=FALSE}
library(knitr)
opts_knit$set(root.dir = "../")
opts_chunk$set(echo = TRUE, message = FALSE, collapse = TRUE, warning = FALSE, cache = TRUE, cache.path = "tut-07-cache/", fig.path = "tut-07-images/")
```


```{r include=FALSE}
library(dplyr)
library(tibble)
library(ggplot2)

# dataset in R, 1973–74 models, Motor Trend
data(mtcars)

# when row names contain data (not tidy)
library(dplyr)
my_mtcars <- mtcars %>%
	mutate(make_model = row.names(mtcars))
```





## condition a dot plot with text

To replace the dot with text, we select a variable in the data frame that has the text we want. 

Here, we'll use *cyl*, the variable that has the number of cylinders in the engine of each car. 

```{r}
# recall that cyl is in the data frame
glimpse(my_mtcars)
```

We use *geom_text* and `label = cyl` in the aesthetics.

```{r}
# could use the number of cylinders as the data marker
f4 <- ggplot(data = my_mtcars
	, aes(x = mpg, y = reorder(make_model, mpg), label = cyl)
	) +
	geom_text()
print(f4)
```

Next tutorial: [condition a dot plot with color](tut-0705_condition-by-color.md)<br>
Tutorial list: [week 2](week-02_assignments.md)  

---
 
[main page](../README.md)<br>
[topics page](../README-by-topic.md) 








