---
output: github_document
bibliography: "../manage/refs.bib" 
csl: "../manage/csl/journal-of-peace-research.csl"
---

```{r echo=FALSE}
library(knitr)
opts_knit$set(root.dir = "../")
opts_chunk$set(echo = TRUE, message = FALSE, collapse = TRUE, warning = FALSE, cache = TRUE, cache.path = "tut-03-cache/", fig.path = "tut-03-images/")
```


```{r echo=FALSE}
library(ggplot2)

time_sec  <- seq(0, 2*pi, by = pi/50) # sec
freq      <- 2                        # r/s
displ_mm  <- sin(freq * time_sec)     # mm
vibr_data <- data.frame(time_sec, displ_mm)

f1 <- ggplot(data = vibr_data, aes(x = time_sec, y = displ_mm)) +
	geom_point()
```

## edit the data markers 

At this point in the tutorial, we expect you to have a *ggplot()* graph object called *f1* that looks like this: 

If not, please return to the earlier tutorials. 

```{r echo=FALSE}
print(f1)
```

Change the data marker using the `shape` argument, taking values between 1 and 25 for the [R default plotting symbols](http://www.cookbook-r.com/Graphs/Shapes_and_line_types/). Shapes numbered 21-25 have both `color` (outline) and `fill` color attributes. 

In the `geom_point()` function, add the argument `shape = 21`.
  
```{r}
f1 <- ggplot(data = vibr_data, aes(x = time_sec, y = displ_mm)) +
	geom_point(shape = 21)

print(f1)
```

We edit the attributes of the data marker further by adding `color`, `fill`, and `size` arguments to `geom_point()`. 

```{r}
f1 <- ggplot(data = vibr_data, aes(x = time_sec, y = displ_mm)) +
	geom_point(shape = 21
		, color = "slateblue4"
		, fill  = "slateblue1"
		, size  = 3
		)
```

In the `geom_point()` argument list, I'm using a "line-break before each comma" style. It may look odd at first, but it has a purpose. I can comment-out a single argument line without affecting the rest of the code. I find that this helps with debugging and reusing code plus makes a long list of arguments easier to read.

```{r}
print(f1)
```

We'll treat color in much more detail later, but for now, the document [Colors in R](http://www.stat.columbia.edu/~tzheng/files/Rcolor.pdf) shows the default R colors by name. 






Next tutorial: [edit axis labels and title](tut-0306_edit-axis-labels-title.md)

---

[main page](../README.md)<br> 
[topics page](../README-by-topic.md)
