---
output: github_document
bibliography: "../manage/refs.bib" 
csl: "../manage/csl/journal-of-peace-research.csl"
---

```{r echo=FALSE}
library(knitr)
opts_knit$set(root.dir = "../")
opts_chunk$set(echo = TRUE, message = FALSE, collapse = TRUE, warning = FALSE, cache.path = "tut-07-images")
```


```{r include=FALSE}
library(dplyr)
library(tibble)
library(ggplot2)

# dataset in R, 1973–74 models, Motor Trend
data(mtcars)

# when row names contain data (not tidy)
my_mtcars <- mtcars %>%
	mutate(make_model = row.names(mtcars))
```


## create a dot plot 

The dot plot can be used to replace any data graphic typically drawn as a pie chart or a bar chart. The data structure is typically 

- a continuous quantitative variable (such as mpg) associated with 
- a categorical variable with a number of levels (such as the car make and model) 
 
 The data frame *my_mtcars* was created in the previous tutorial. 
 
 Here we graph the car make and model as a function of mileage in mpg. 
 
```{r}
# make basic dot plot
library(ggplot2)
f1 <- ggplot(data = my_mtcars, aes(x = mpg, y = make_model)) +
	geom_point()
print(f1)
```

Next tutorial: [reorder the rows of a dot plot](tut-0703_reorder-rows.md)<br>
Tutorial list: [week 2](week-02_assignments.md)  

---
 
[main page](../README.md)<br>
[topics page](../README-by-topic.md)  



