---
output: github_document
---

```{r echo=FALSE}
library(knitr)
opts_knit$set(root.dir = "../")
opts_chunk$set(echo = TRUE, messages = FALSE, collapse = TRUE)
```

## starting a practiceR script 

Start every session for our course work by launching your  *me447_visualizing-data.Rproj* file, 

*File* $\rightarrow$ *Open Project* $\rightarrow$ *me447_visualizing-data.Rproj*

Using an R Project sets the working directory to your course folder, making relative file paths easy to use.

## scatterplot 

Start a new R script and save it to your *practiceR* directory. 

- *File* $\rightarrow$ *New File* $\rightarrow$ *R Script*
- *File* $\rightarrow$ *Save As ...* $\rightarrow$ **tut02_scatterplot.R*

My suggested file naming 

- starts with *tut* indicating a tutorial  
- follows with with *02* indicating chronological order in which the tuitorials are assigned 
- concludes with *scatterplot* indicating the topic 

Begin the file with a header, something like this. 

```
#  ------------------------------------------
#  start learning ggplot2 using a scatterplot
#
#  Richard Layton
#  2017-08-29
#  ------------------------------------------
```

Clearing variables from your workspace is optional.

```{r}
# if you like to start with an empty workspace 
rm(list = (ls()))
```

Style notes: 

- Place spaces around all operators (`=, +, -, <-, *`, etc.).  
- Comments should explain the why, not the what. 
- See Hadley Wickham's [Advanced R](http://adv-r.had.co.nz/Style.html) for more on R coding style guidelines. (Hadley is "Internet famous" in the R community for creating marvelous packages.)

To run the script after adding a new code chunk, 

- Press the *Source* button (in the upper-right corner of the script pane) to run your script. 
- I recommend you acquire this habit to help you identify and debug problems as they occur.  
- After running a script, the Console pane will show a prompt and a cursor,  `> |` indicating that your script executed with no errors. 


## course directory check 

We added new files to your directory

    me447_visualizing-data
      |- data
      |- design 
      |- portfolio 
      |- practiceR 
      |		|- tut01_first-script.R
      |		`- tut02_scatterplot.R
      |- visuals
      `- me447_visualizing-data.Rproj


--- 

Next tutorial: [install and load a package](tut-0302_install-load-package.md) 

---

[main page](../README.md)<br> 
[topics page](README-by-topic.md)
