---
output: github_document
bibliography: "../manage/refs.bib" 
csl: "../manage/csl/journal-of-peace-research.csl"
---

```{r echo=FALSE}
library(knitr)
opts_knit$set(root.dir = "../")
opts_chunk$set(echo = TRUE, message = FALSE, collapse = TRUE, warning = FALSE, cache = TRUE, cache.path = "tut-04-cache/", fig.path = "tut-04-images/")
```

```{r echo=FALSE}
rm(list = (ls()))
library(tidyverse)
```

## accessing data in R 

To see the data sets included in your R workspace, use the *data()* function. 

```{r}
# accessing data in R
data()
```

A tab will open in the RStudio pane that looks like this: 

```
Data sets in package 'datasets':

AirPassengers      Monthly Airline Passenger Numbers 1949-1960
BJsales            Sales Data with Leading Indicator
BOD                Biochemical Oxygen Demand
CO2                Carbon Dioxide Uptake in Grass Plants
ChickWeight        Weight versus age of chicks on different diets
```

You can scroll down this list looking for interesting data.  Some of these datasets are used regularly by R folks to illustrate package functionality. 

- Motor Trend car data. Type `?mtcars` in the Console to read more about it. 

```{r}
# a commonly used dataset for R examples
glimpse(mtcars)
```

- Three species of iris (the flower). Type `?iris` in the Console to read more about it. 

```{r}
# a commonly used dataset for R examples
glimpse(iris)
```

- Survival of passengers on the Titanic. Type `?Titanic` in the Console to read more about it. 

```{r}
# a commonly used dataset for R examples
glimpse(Titanic)
```

- Star Wars characters, from the *dplyr* package. Type `?starwars` in the Console to read more about it. 

```{r}
# a commonly used dataset for R examples
glimpse(starwars)
```

- US economic time series, from the *ggplot2* package. Type `?economics` in the Console to read more about it. 

```{r}
# a commonly used dataset for R examples
glimpse(economics)
```

- World Health Organization tuberculosis data, from the *tidyr* package. Type `?who` in the Console to read more about it. 

```{r}
# a commonly used dataset for R examples
glimpse(who)
```


Next tutorial: [data basics exercise](tut-0408_data-basics-exercise.md)

---

[main page](../README.md)<br> 
[topics page](../README-by-topic.md)


