---
output: github_document
bibliography: "../manage/refs.bib" 
csl: "../manage/csl/journal-of-peace-research.csl"
---

```{r echo=FALSE}
library(knitr)
opts_knit$set(root.dir = "../")
opts_chunk$set(echo = TRUE, message = FALSE, collapse = TRUE, warning = FALSE, cache = TRUE, cache.path = "tut-03-cache/04-", fig.path = "tut-03-images/04-")
```


```{r echo=FALSE}
library(ggplot2)

time_sec  <- seq(0, 2*pi, by = pi/50) # sec
freq      <- 2                        # r/s
displ_mm  <- sin(freq * time_sec)     # mm
vibr_data <- data.frame(time_sec, displ_mm)
```

## create a scatterplot 

At this point in the tutorial, we expect you to have a data frame *vibr_data* in your workspace with columns *time_sec* and *displ_mm*. If not, please return to the earlier tutorials. 

The structure of our *ggplot2* scatterplot begins with identifying the data frame, the columns assigned as *x* and *y* variables, and using *points* as data markers.   

```{r}
# basic scatterplot
f1 <- ggplot(data = vibr_data, aes(x = time_sec, y = displ_mm)) +
	geom_point()

print(f1)
```

What the commands mean

- `ggplot()` initializes a ggplot object
- `data = ` assigns the data frame with our data 
- `aes()` describes the "aesthetic mapping" from data to visual properties
- `geom_point()` draws points on a scatterplot 
- `print()` prints the graph onscreen in the RStudio *Plots* pane. 









Next tutorial: [edit data markers](tut-0305_edit-data-markers.md)


---

[main page](../README.md)<br> 
[topics page](../README-by-topic.md)
