---
output: github_document
bibliography: "../manage/refs.bib" 
csl: "../manage/csl/journal-of-peace-research.csl"
---

```{r echo=FALSE}
library(knitr)
opts_knit$set(root.dir = "../")
opts_chunk$set(echo = TRUE, message = FALSE, collapse = TRUE, warning = FALSE, cache.path = "tut-07-images")
```



## capture data stored in row names

We'll start with some (nearly) tidy data. 

```{r}
# dataset in R, 1973–74 models, Motor Trend
data(mtcars)
head(mtcars)
```

This data set has information stored in the row names of the data frame. So, not quite tidy. 

The *dplyr* package (for data manipulation) gives us the "pipe" operator `%>%`  and  *mutate()* that creates a new variable (column) in the data frame. 

```{r}
# copy the row names to a new variable "make_model" 
library(dplyr)
my_mtcars <- mtcars %>%
	mutate(make_model = row.names(mtcars))
```

Examine the result. You can see that a new variable *make_model* has been added to the data frame. 

```{r}
class(my_mtcars)

glimpse(my_mtcars)
```

Next tutorial: [create a dot plot](tut-0702_create-dot-plot.md)<br>
Tutorial list: [week 2](week-02_assignments.md)  

---
 
[main page](../README.md)<br>
[topics page](../README-by-topic.md)  






