---
title: tables
output: 
  github_document:
    toc: TRUE
    toc_depth: 2
bibliography: "../manage/refs.bib" 
csl: "../manage/csl/journal-of-peace-research.csl"
---

```{r 1301, echo=FALSE}
library(knitr)
opts_knit$set(root.dir = "../")
opts_chunk$set(echo = TRUE, message = FALSE, collapse = TRUE, warning = FALSE, cache = TRUE, cache.path = "tut-13-cache/01-", fig.path = "tut-13-images/01-")
```


```{r 1302, echo = FALSE}
library(tidyverse)
```

## preparation

- Open a new Rmd script called *tut13_tables.Rmd*
- Save it to your *practiceR* folder 
- Edit YAML header 
- Delete the remaining default contents 

Add the usual knitr code chunk 

```
library(knitr)
opts_knit$set(root.dir = "../")
opts_chunk$set(echo = TRUE, messages = FALSE)
```

Load the libraries (you may have to install *pander* first.)

```
library(tidyverse)
library(pander)
```

Knit to Word. 




## tabulate some data

```{r}
# create a subset of the mtcars data
df <- mtcars %>%
  mutate(make_model = row.names(mtcars)) %>%
  filter(near(cyl, 4)) %>%    
  select(make_model, mpg, wt) %>%
  mutate(wt = wt*1000) %>%
  arrange(desc(wt)) 
```

The simplest approach to printing a table is to just print the data frame as-is. Not particularly readable, but easy. 

Add this sentence to your Rmd as prose. 

> Here is a data frame printed with no additional formatting. 

Then add this code chunk. 

```{r eval=FALSE}
# just the data frame as is
print(df)
```

The output:

```{r echo=FALSE}
print(df)
```

Knit to Word. 




## tables with *kable()*  

The next approach is to use the *knitr* function *kable()*. We've used it before. First type a new sentence into your Rmd script, for example, 

> Here is a paragraph just before a table by kable(). 

Add the code chunk. 

```{r eval=FALSE}
# use knitr::kable() to print a data frame as a table 
kable(df) 
```

The output: 

```{r echo=FALSE}
kable(df) 
```

Knit to Word. 

The default alignment agrees with our basic principles. For data shared within your working group, *kable()* is pretty good. For professional publication, we'll need something else. 



## use `<br>` for an extra empty line  

I notice that I need some white space to separate text from the code chunks and from the table. We can use the HTML line break tag `<br>` on its own line to create the extra white space.

Add the `<br>` markup on its own line, followed by a blank line before and after the sentence "Here is a paragraph just before a table by kable()." 

Your markup in Rmd should look like this


    <br>
  
    Here is a paragraph just before a table.
  
    <br>


Knit to Word. 

You should see that the `<br>` markup placed an extra line break on either side of he prose paragraph, giving you some additional white space before the table. I use this regularly when I want a single extra line. 





## tables with *pander()* 

 I usually place initializing code chunks like this at the top of a script. The *right* and *left* arguments in the function will align text data left and numeric data right. 

```{r}
# set pander table-layout options
library(pander)
panderOptions('table.alignment.default', function(df)
    ifelse(sapply(df, is.numeric), 'right', 'left'))
panderOptions('table.split.table', Inf)
panderOptions('big.mark', ",")
panderOptions('keep.trailing.zeros', TRUE)
```

Type a new sentence into your Rmd script, for example, 

> Here is a paragraph just before a table by pander(). 

Add the code chunk. 

```{r eval=FALSE}
# use pander() to print a data frame as a table 
pander(df) 
```

The output: 

```{r echo=FALSE}
pander(df) 
```

Knit to Word.  What we gained here is the ability to

- the thousands separator (`big.mark`) is a comma 
- to keep training zeros 
- and still align text left and numbers right  





## auto-numbering tables

As Norbert K&ouml;hler says, "R Markdown has no native method to number and reference table and figure captions". The `captioner` packages gives us a method  for auto-numbering figures and tables and automatically updating the in-text reference numbers  as well. 

This tutorial is adapted from Norbert K&ouml;hler's [blog post](http://datascienceplus.com/r-markdown-how-to-number-and-reference-tables/) plus Andrew Dolman's [comments](https://onedrive.live.com/?authkey=%21AApYWBGHNR06Y5I&cid=FCFCD3FD042AF7F1&id=FCFCD3FD042AF7F1%2118306&parId=FCFCD3FD042AF7F1%21322&action=locate) on that post. 


When complete, your output document should have two additional tables with automatically-generated table numbers. 

- Install the `captioner` package 

Add the code chunk to your Rmd script

```{r}
library(captioner)
```

Edit the output portion of the YAML header. 

```
output:
  word_document:
    fig_caption: yes
    reference_docx: '../resources/word-styles-reference-01.docx'
```

### assign prefixes

Use `captioner()` to set the prefixes we want to use for our tables and figures. Add this code chunk to your Rmd script. 

```{r results = 'hide'}
# assign the prefix
auto_tab_ID <- captioner(prefix = "Table")
```

The function names on the left, `auto_tab_ID` and `auto_fig_ID`, are arbitrary. 

We create a table captions using the `auto_tab_ID()` function we created above. The `name` argument is a unique identification string that we assign and the `caption` argument is a string with your actual caption. 

```{r results = 'hide'}
# assign the table prefix, number, and caption 
auto_tab_ID(name = "a_table", caption = "A descriptive caption for the table.")
```

In the code chunk above---add `include = FALSE` to the code chunk header. 


### refer to the table inline  

In a typical technical report, the paragraph just before a table includes an inline reference to the table. For example, add this prose to your Rmd script, 

<pre><code>An excerpt of data from our fabulous experiment is shown in 
<code>`</code>r auto_tab_ID("a_table", display = "cite")<code>`</code>. 
</code></pre>

Knit Word and you should see that "Table 1" has been inserted in your sentence for you. 

The single tick marks delineate an inline R code chunk; the `auto_tab()` function uses the identifier string we created above and will print the caption with a table number. The argument `display = "cite"` prints the table number inline but not the caption. 

### print the table with caption 

The next code chunk, using kable(), prints the Table number, the caption, and the table. 

```{r eval = FALSE}
# print the table with caption
kable(df, caption = auto_tab_ID("a_table"))
```

In docx output, table captions are formatted with the *Table Caption* style. The style can be edited using the methods for controlling Word Styles we've used earlier. 



### make a second table 

I'm going to create another table to demonstrate that the next table is indeed  assigned the number "2". 

```{r results = 'hide'}
# create a caption for a second table 
auto_tab_ID("another_table", "A boring table.")
```

Having created another caption, I can make an inline reference to the next table.

> More amazing results are shown in `r auto_tab_ID("another_table", display = "cite")`. 

Created using this Rmd script: 

<pre><code>More amazing results are shown in 
<code>`</code>r auto_tab_ID("another_table", display = "cite")<code>`</code>. 
</code></pre>

New table: 

```{r eval=FALSE}
# get some data
df2 <- tail(cars, n = 5L)

# print the table with caption
kable(df2, caption = auto_tab_ID("another_table"), row.names = FALSE)
```

All tables are automatically renumbered if tables are added or deleted. 




















