---
output: github_document
bibliography: "../manage/refs.bib" 
csl: "../manage/csl/journal-of-peace-research.csl"
---

```{r echo=FALSE}
library(knitr)
opts_knit$set(root.dir = "../")
opts_chunk$set(echo = TRUE, message = FALSE, collapse = TRUE, warning = FALSE, cache = TRUE, cache.path = "tut-03-cache/09-", fig.path = "tut-03-images/09-")
```

## scatterplot exercise

`ggplot2` includes a data frame named `diamonds`. In the console, type `?diamonds` to read a description. 

Use the methods presented in this tutorial to answer these questions. 

1. How many diamonds were measured?  
2. How many variables in the data set are numerical?  
2. Create a scatterplot to illustrate the potential dependence of price on carats.  
3. Edit the following to obtain a graph you find satisfactory 
    - data marker size, shape, and color 
    - axis labels and title
    - theme
    - axis and title text size and type 

What visual problem have we not yet addressed? 

```{r echo = FALSE, eval = FALSE}
str(diamonds)

ggplot(diamonds, aes(x = carat, y = price)) +
  geom_point(shape = 16, size = 0.5)
```





---

[main page](../README.md)<br> 
[topics page](../README-by-topic.md)
