---
output: github_document
bibliography: "../manage/refs.bib" 
csl: "../manage/csl/journal-of-peace-research.csl"
---

```{r echo=FALSE}
library(knitr)
opts_knit$set(root.dir = "../")
opts_chunk$set(echo=TRUE, message=FALSE, collapse=TRUE, warning=FALSE)
```

```{r include=FALSE}
rm(list = (ls()))
detachAllPackages <- function() {

  basic.packages <- c("package:stats","package:graphics","package:grDevices","package:utils","package:datasets","package:methods","package:base")

  package.list <- search()[ifelse(unlist(gregexpr("package:",search()))==1,TRUE,FALSE)]

  package.list <- setdiff(package.list,basic.packages)

  if (length(package.list)>0)  for (package in package.list) detach(package, character.only=TRUE)

}
detachAllPackages()
library(tidyverse)
library(readxl)
library(rmarkdown)
search()
```


## when a new library masks a function

In the previous tutorial, we loaded the *VIM* package and a message was printed indicating that an object is "masked." The message indicates that the *VIM* package defines functions *between()*, *first()*, and *last()* that are also functions in the *dplyr* package. Because *VIM* was loaded last, its functions "mask" the *dplyr* functions of the same name. 


```{r message=TRUE}
# load the package
library(VIM)
```

This issue arises because people are constantly writing new packages and there is inevitably some overlap in naming functions. 

Since I use *dplyr* a lot but *VIM* only ocassionally, I can "unload" *VIM* once I'm done with it using the *unloadNamespace()* function: the opposite of the *library()* function. 

```{r}
# unload a package
unloadNamespace("VIM")
```

Use *search()* to confirm that *VIM* is no longer loaded. 

```{r}
# list packages currently loaded
search()
```







Next tutorial: [add a regression line to a plot](tut-0405_add-regression.md)

---

[main page](../README.md)<br> 
[topics page](../README-by-topic.md)



