---
output: github_document
bibliography: "../manage/refs.bib" 
csl: "../manage/csl/journal-of-peace-research.csl"
---

```{r echo=FALSE}
library(knitr)
opts_knit$set(root.dir = "../")
opts_chunk$set(echo = TRUE, message = FALSE, collapse = TRUE, warning = FALSE, cache = TRUE, cache.path = "tut-07-cache/", fig.path = "tut-07-images/")
```



```{r include=FALSE}
library(dplyr)
library(tibble)
library(ggplot2)

# dataset in R, 1973–74 models, Motor Trend
data(mtcars)

# when row names contain data (not tidy)
library(dplyr)
my_mtcars <- mtcars %>%
	mutate(make_model = row.names(mtcars))
library(tibble)

# OK, but treated as continuous var, actually discrete
my_mtcars <- my_mtcars %>%
	mutate(Cylinders = factor(cyl))

# instead of colors, could use panels
f8 <- ggplot(data = my_mtcars, aes(x = mpg, y = reorder(make_model, mpg))) +
	geom_point(size = 3) +
	facet_grid(~ Cylinders)
```






## set the y scales free 

An argument added to *facet_grid()*. 

Because each car has only one mpg, when we used *facet_grid()*, we end up with empty rows in each panel.

To eliminate the unused rows, first we change panel orientation to vertical. 

```{r}
# orient vertically
f10 <- f8 +
	facet_grid(Cylinders ~.)
print(f10)
```

Next use `scales = "free_y"` to eliminate all rows with no *mpg* data. 

```{r}
# free the scales to remove lines with no data
f11 <- f8 +
	facet_grid(Cylinders ~., scales = "free_y")
print(f11)
```


To allow the panel heights to vary such that the row spacing is consistent, add `space = "free_y"` to *facet_grid()*. 

```{r}
# make the spacing free too
f12 <- f8 +
	facet_grid(Cylinders ~., scales = "free_y", space = "free_y")
print(f12)
```


Next tutorial: [rotate text in the panel strip](tut-0710_rotate-strip-text.md)<br>
Tutorial list: [week 2](week-02_assignments.md)  

---
 
[main page](../README.md)<br>
[topics page](../README-by-topic.md) 



