---
title: displaying univariate data
output: 
  github_document:
    toc: TRUE
    toc_depth: 2
bibliography: "../manage/refs.bib" 
csl: "../manage/csl/journal-of-peace-research.csl"
---

```{r 1501, echo=FALSE}
library(knitr)
opts_knit$set(root.dir = "../")
opts_chunk$set(echo = TRUE, message = FALSE, collapse = TRUE, warning = FALSE, cache = TRUE, cache.path = "tut-15-cache/01-", fig.path = "tut-15-images/01-")
```


```{r 1502, echo = FALSE}
library(tidyverse)
```

## preparation

Start by opening your RStudio project:    
File $\rightarrow$ Recent projects $\rightarrow$ me447_visualizing-data.Rproj.

> ALWAYS start your work session by launching the .Rproj file you created for this course 

Using an R Project sets the working directory to your course folder, making relative file paths easy to use.

- Open a new R script called *tut15_univariate.R*
- Save it to your *practiceR* folder 

Add the usual knitr code chunk 

```
library(knitr)
opts_knit$set(root.dir = "../")
opts_chunk$set(echo = TRUE, messages = FALSE)
```

Load packages.


```
library(tidyverse)
```


- Download the xlsx file *univariate_doumont_cases.xlsx* from the *data* directory on GitHub. 
- Save to your *data* directory. 








## bibliography 

<div id="refs"></div>
  
---

[main page](../README.md)<br>
[topics page](../README-by-topic.md)
