---
output: github_document
bibliography: "../manage/refs.bib" 
csl: "../manage/csl/journal-of-peace-research.csl"
---

```{r echo=FALSE}
library(knitr)
opts_knit$set(root.dir = "../")
opts_chunk$set(echo = TRUE, message = FALSE, collapse = TRUE, warning = FALSE, cache = TRUE, cache.path = "tut-03-cache/06-", fig.path = "tut-03-images/06-")
```


```{r echo=FALSE}
library(ggplot2)

time_sec  <- seq(0, 2*pi, by = pi/50) # sec
freq      <- 2                        # r/s
displ_mm  <- sin(freq * time_sec)     # mm
vibr_data <- data.frame(time_sec, displ_mm)

f1 <- ggplot(data = vibr_data, aes(x = time_sec, y = displ_mm)) +
	geom_point(shape = 21
		, color = "slateblue4"
		, fill  = "slateblue1"
		, size  = 3
		)
```

## edit axis labels and title 

At this point in the tutorial, we expect you to have a *ggplot()* graph object called *f1* that looks like this: 

If not, please return to the earlier tutorials. 

```{r echo=FALSE}
print(f1)
```

Label the axes, replacing the default axis labels. 

```{r}
# add axis labels
f1 <- f1 + 
	xlab("Time (sec)") + 
	ylab("Position (mm)")
print(f1)
```

Add a main title. 
 
```{r}
# add a title
f1 <- f1 + 
	ggtitle("Scatterplot")
print(f1)
```

I'm showing you one or two lines of code at a time so you can add the lines, run the script, and see the results one chunk at a time. 

In practice, I usually just keep adding lines to the original `ggplot()` function using the plus (+) function, like this.   

```
# chaining functions together using the plus (+) sign
f1 <- ggplot(data = vibr_data, aes(x = time_sec, y = displ_mm)) +
	geom_point(shape = 21
		, color = "slateblue4"
		, fill  = "slateblue1"
		, size  = 3
		) +
	xlab("Time (sec)") + 
	ylab("Position (mm)") +
	ggtitle("Scatterplot")
	
print(f1)
```




Next tutorial : [edit the scales](tut-0307_edit-scales.md) 

---

[main page](../README.md)<br> 
[topics page]../README-by-topic.md)
