---
output: github_document
bibliography: "../manage/refs.bib" 
csl: "../manage/csl/journal-of-peace-research.csl"
---

```{r echo=FALSE}
library(knitr)
opts_knit$set(root.dir = "../")
opts_chunk$set(echo=TRUE, message=FALSE, collapse=TRUE, warning=FALSE)
```

```{r echo=FALSE}
library(tidyverse)
```

## data basics exercises

- Install the *fivethirtyeight* package and view its help page.  
- Load the  *fandango* data set.

1. Movie ratings 
    - Draw a scatterlot showing the `rottentomatoes` ratings as a function of the `metacritic` ratings. 
    - Draw a second scatterplot showing the `fandango_ratingvalue` as a function of the `metacritic` data. 
    - What conclusions do you draw? 
2. Missing values 
    - Use VIM to determine if *fandango* has missing values. 

```{r include=FALSE}
library(fivethirtyeight)
glimpse(fandango)

library(VIM)
aggr_plot <- aggr(
	fandango
	, numbers  = TRUE
	, sortVars = TRUE
	, prop     = FALSE
	, combined = TRUE
	, ylab     = "Pattern of missing values"
)
unloadNamespace("VIM")
```


---

[main page](../README.md)<br> 
[topics page](../README-by-topic.md)


