---
output: github_document
---

```{r echo=FALSE}
library(knitr)
opts_knit$set(root.dir = "../")
opts_chunk$set(echo = TRUE, messages = FALSE, collapse = TRUE)
```

## edit the scales

Our scatterplot script to this point consists of

```{r}
library(ggplot2)

time_sec  <- seq(0, 2*pi, by = pi/50) # sec
freq      <- 2                        # r/s
displ_mm  <- sin(freq * time_sec)     # mm
vibr_data <- data.frame(time_sec, displ_mm)

f1 <- ggplot(data = vibr_data, aes(x = time_sec, y = displ_mm)) +
	geom_point(shape = 21
		, color = "slateblue4"
		, fill  = "slateblue1"
		, size  = 3
		) +
	xlab("Time (sec)") + 
	ylab("Position (mm)") +
	ggtitle("Scatterplot")
```

Suppose we want the graph to have a 2 mm y-scale with tick marks at 1 mm intervals. 

```{r}
f1 <- f1 +
	scale_y_continuous(limits = c(-2, 2), breaks = seq(-2, 2, 1))

print(f1)
```

---

Self-check quiz

1. Explain the values assigned to the `limits` argument  
2. Explain the values assigned to the `breaks` argument  

Confirm your answers by typing the following in your Console and reading the results in *Help* pane.  

- `?c()` 
- `?scale_y_continuous`


---





Previous tutorial: [edit axis labels and title](tut-0306_edit-axis-labels-title.md)<br>
Next tutorial: [edit theme](tut-0308_edit-theme.md)

---

[main page](../README.md)<br> 
[topics page](README-by-topic.md)
