---
output: github_document
bibliography: "../manage/refs.bib" 
csl: "../manage/csl/journal-of-peace-research.csl"
---

```{r echo=FALSE}
library(knitr)
opts_knit$set(root.dir = "../")
opts_chunk$set(echo = TRUE, message = FALSE, collapse = TRUE, warning = FALSE, cache.path = "tut-07-images")
```


```{r include=FALSE}
library(dplyr)
library(tibble)
library(ggplot2)

# dataset in R, 1973–74 models, Motor Trend
data(mtcars)

# when row names contain data (not tidy)
library(dplyr)
my_mtcars <- mtcars %>%
	mutate(make_model = row.names(mtcars))
```






## condition a dot plot with color

We use *geom_point* and `col = cyl` in the aesthetics.

```{r}
# could use dot color
f5 <- ggplot(data = my_mtcars
	, aes(x = mpg, y = reorder(make_model, mpg), col = cyl)
  ) +
	geom_point(size = 3)
print(f5)
```

The legend indicates color-coding used for a continuous variable, but the number of cylinders is actually a discrete value with only 3 levels: 4, 6, and 8.

We use *factor()* to change cylinders from a double-precision variable to a factor and *mutate()*  to create the new variable in the data frame. 

```{r}
# new variable so color can be assigned discrete
my_mtcars <- my_mtcars %>%
	mutate(Cylinders = factor(cyl))
```

We change the color aesthetic to use the new variable, `col = Cylinders`. 

```{r}
# again with factor
f6 <- ggplot(data = my_mtcars
	, aes(x = mpg, y = reorder(make_model, mpg), col = Cylinders)
	) +
	geom_point(size = 3)
print(f6)
```


Next tutorial: [manually change dot plot colors](tut-0706_manual-color.md)<br>
Tutorial list: [week 2](week-02_assignments.md)  

---
 
[main page](../README.md)<br>
[topics page](../README-by-topic.md) 






