---
title: install and load packages
output: 
  github_document:
    toc: TRUE
    toc_depth: 2
bibliography: "../manage/refs.bib" 
csl: "../manage/csl/journal-of-peace-research.csl"
---

```{r echo=FALSE}
library(knitr)
opts_knit$set(root.dir = "../")
opts_chunk$set(echo = TRUE, message = FALSE, collapse = TRUE, warning = FALSE, cache = TRUE, cache.path = "tut-03-cache/02-", fig.path = "tut-03-images/02-")
```

## preparation

*Packages* are R modules that you can download and install (like apps for R). Before installing any packages, make sure have done the following:

- Did you start this session by launching your .Rproj file for the course? If not, select File $\rightarrow$ Recent projects $\rightarrow$ me447_visualizing-data.Rproj. 
- Does your top-level directory for the course include the .Renviron file? If not, revisit the [create top level package library](https://github.com/DSR-RHIT/install-R-and-RStudio/blob/master/pages/setup_Renviron.md) page. 

## install and load a package

Watch this [1-minute video](https://www.youtube.com/watch?v=ljdfqMfWn_A) to learn how to install an R package. 

- Next, install the `tidyverse` package (by Hadley Wickham). This package encompasses an entire suite of useful packages, including `ggplot2`. 

Once a package is installed, you load the package using the `library()` command.

```{r}
# to access the functionality of the ggplot2 package 
library(ggplot2)
```

Again, add this line of code to your script and run the script by pressing the *Source* button. 

Next tutorial: [create a data frame](tut-0303_create-data-frame.md) 

---

[main page](../README.md)<br> 
[topics page](../README-by-topic.md)
