---
output: github_document
bibliography: "../manage/refs.bib" 
csl: "../manage/csl/journal-of-peace-research.csl"
---

```{r echo=FALSE}
library(knitr)
opts_knit$set(root.dir = "../")
opts_chunk$set(echo = TRUE, message = FALSE, collapse = TRUE, warning = FALSE, cache.path = "tut-07-images")
```


```{r include=FALSE}
library(dplyr)
library(tibble)
library(ggplot2)

# dataset in R, 1973–74 models, Motor Trend
data(mtcars)

# when row names contain data (not tidy)
my_mtcars <- mtcars %>%
	mutate(make_model = row.names(mtcars))
```




## reorder the rows of a dot plot

Apply *reorder()* to the y-variable and indicate the numerical variable used to order the rows. 

```{r}
# reorder rows
f2 <- ggplot(data = my_mtcars, aes(x = mpg, y = reorder(make_model, mpg))) +
	geom_point()
print(f2)
```

Let's re-size the dots too. 

```{r}
# dots bigger
f3 <- ggplot(data = my_mtcars, aes(x = mpg, y = reorder(make_model, mpg))) +
	geom_point(size = 3)
print(f3)
```

That's pretty much it for a single-panel, basic dot plot. Next we investigate different ways to use conditioning variables. 

Next tutorial: [condition a dot plot with text](tut-0704_condition-by-text.md)<br>
Tutorial list: [week 2](week-02_assignments.md)  

---
 
[main page](../README.md)<br>
[topics page](../README-by-topic.md) 
