---
output: github_document
bibliography: "../manage/refs.bib" 
csl: "../manage/csl/journal-of-peace-research.csl"
---

```{r echo=FALSE}
library(knitr)
opts_knit$set(root.dir = "../")
opts_chunk$set(echo = TRUE, message = FALSE, collapse = TRUE, warning = FALSE, cache = TRUE, cache.path = "tut-03-cache/08-", fig.path = "tut-03-images/08-")
```


```{r echo=FALSE}
library(ggplot2)

time_sec  <- seq(0, 2*pi, by = pi/50) # sec
freq      <- 2                        # r/s
displ_mm  <- sin(freq * time_sec)     # mm
vibr_data <- data.frame(time_sec, displ_mm)

f1 <- ggplot(data = vibr_data, aes(x = time_sec, y = displ_mm)) +
	geom_point(shape = 21
		, color = "slateblue4"
		, fill  = "slateblue1"
		, size  = 3
		) +
	xlab("Time (sec)") + 
	ylab("Position (mm)") +
	ggtitle("Scatterplot") +
	scale_y_continuous(limits = c(-2, 2), breaks = seq(-2, 2, 1))
```

## edit the plot theme

At this point in the tutorial, we expect you to have a *ggplot()* graph object called *f1* that looks like this: 

If not, please return to the earlier tutorials. 

```{r echo=FALSE}
print(f1)
```


The gray background is a *ggplot2* default setting. We can use ggplot2 *themes* to change the background (and every other default setting as well). 

In the first example, we'll use one of the set of themes available in the *ggplot2* package. 

```{r}
f1 <- f1 +
	theme_light()

print(f1)
```

Examples of all the default themes are illustrated at Wickham's [ggplot2 website](http://ggplot2.tidyverse.org/reference/ggtheme.html).  

Control the size and font style using the `theme()` function.  I use the `rel()` size argument to make all text the same relative size (`0.9`) and type (`"plain"`). 

```{r}
f1 <- f1 + 
	theme(plot.title = element_text(size = rel(0.9), face = "plain")  
		, axis.title = element_text(size = rel(0.9), face = "plain") 
		, axis.text = element_text(size = rel(0.9), face = "plain")
		)

print(f1)
```




Next tutorial: [scatterplot exercise](tut-0309_scatterplot-exercise.md) 

---

[main page](../README.md)<br> 
[topics page](../README-by-topic.md)
